---
title: "Lab 4"
author: "brouwern@gmail.com"
date: "September 21, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Task 1: Data Entry
Enter the lengths of each your  genes into a **vector** using the **c( )** command.

```{r}
my.genes <- c(1042,1472,4339,1865,4706,2439,3369,2988,3286, 3773)
```

# Task 2: Summary Statistics

### 2a) Confirm the sample size with the **length()** command
```{r}
length(my.genes)
```


### 2b) Calculate the median, the min, and the max using the appropriate R commands
```{r}
median(my.genes)
min(my.genes)
max(my.genes)
```



### 2c) Calculate the mean "long hand" using the **sum()** and **length()** command.  Enter this into your worksheet.  Check your answer using the appropriate R command.
```{r}
sum(my.genes)/length(my.genes)
```

### 2d) Calculate the variance using the **var()** command
```{r}
var(my.genes)
```

### 2e) Calculate the standard deviation (sd) "long hand" using the square root of the variance.

\[\sigma = \sqrt{\frac{\sum\limits_{i=1}^{n} \left(x_{i} - \bar{x}\right)^{2}} {n-1}}\]

```{r}
sqrt(var(my.genes))
```



### 2f) Calculate the standard error (se) "long hand" using the square root of the variance and the length of the data.  The required commands are 

* sqrt()
* var() 
* length()

```{r}
sqrt(var(my.genes))/sqrt(length(my.genes))
```




