---
title: "Intro to t-test"
author: "brouwern@gmail.com"
date: "September 29, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preliminaries
Set working directory
```{r}
setwd("C:/Users/lisanjie2/Desktop/TEACHING/1_STATS_CalU/1_STAT_CalU_2016_by_NLB/Lab/Lab5_intro_to_t_test_and_hypothesis_testing")
```


# Load Brook Trout Data
These data are from Example 12.4 "So long; thanks to all the fish" in Whitlock and Shulter 2nd ed.
```{r}
salmon <- read.csv("Lab5_data_brook_trout.csv")
```


# Examine the brook trout data
Each line of data is a stream that either has introduced brook trout present or not present.
```{r}
# size of dataframe
dim(salmon)

#top of data
head(salmon)

#bottom of data
tail(salmon)

#summary of data
summary(salmon)
```


# Data setup
The **response variable** in this study is the **percent (%)** of chinook salmon that survived over the course of the study.  We therefore need to calcualte this value using R's basic math function for divison.


## Calcualte survival rates

### 1st, calcualte the fraction of fish that surved in each stream
```{r}
percent.surv <-salmon$salmon.surv/salmon$salmon.released
```


### 2nd, check to make sure that the data make sense.  
The data are a percentage, so each datum should be between 0 and 1.  We can look at the data using **summary()**
```{r}
summary(percent.surv)
```

We could also use range(), min(), and max(). to get this info.  We could also make a histogram using hist().


### 3rd, add the percentages to the original dataframe

We'll call the new column "percent.surv" also.
```{r}
salmon$percent.surv <- percent.surv
```
Note: this isn't actually a percent, but a fraction.  To get a percent we'd multiple it by 100.



### 4th, check the calculated data
```{r}
#The whole datafame
salmon

#summary()
summary(salmon)
```



# Data exploration

We should explore the data using boxplots and histograms.  The **response variable** is "percent.surv" and the **predictor variable* is whether introduced brook trout are present or absent.


## Boxplot
```{r}
boxplot(percent.surv~brook.trout.PRES.ABS,
        data = salmon)
```

Median survival is a bit lower but there is not much difference.


## Histograms
Boxplots are adequate for looking at these data, but let's practie make a 2-panel histrogram for practic.  This requires using the **subset()** command to split the data up by group




### Subset the data 
Subset (split) from just the streams where brookies are present.
```{r}
present <- subset(salmon,
                  select = c("brook.trout.PRES.ABS",
                             "percent.surv"),         
                  brook.trout.PRES.ABS == "present")

```


Subset from just the streams where brookies are absetn
```{r}
absent <- subset(salmon,
                  select = c("brook.trout.PRES.ABS",
                             "percent.surv"),         
                  brook.trout.PRES.ABS == "absent")

```



### Make side by side histograms

First, we need to set up the plot window using the **par()** command.  Then we can make the histograms.  We specify the subset dataframe to use (present or absent) and the column within the dataframe **$percent.surv**
```{r}
# the par() command
par(mfrow = c(1,2))

#left histogram
hist(present$percent.surv)

#right histogram
hist(absent$percent.surv)
```




### Make the histograms nicer to look at

* R always makes big titles by default.  We can fix the by including the arguement **main = "..."**, where **main** specifies the main title of the plot.  
* R also will make the x axis different on each plot.  We can fix that using the **xlim** arguement.  THe data are percentages so its natural use use 0 to 1 as the limits.  However, since there are no values above 0.5 we'll use that as the upper limit.
* We can also label the x axis using **xlab = "..."***
* See this page for more info on setting plot limits: https://rpubs.com/brouwern/211902

```{r}
par(mfrow = c(1,2))
#left histogram
hist(present$percent.surv,
     main = "Brk Trout Present", 
     xlim = c(0,0.5),
     xlab = "Percent surv")

#right histogram
hist(absent$percent.surv,
     main = "Brk Trout Absent" ,
     xlim = c(0,0.5),
     xlab = "Percent surv")

```


# Calculate means & stand dev using tapply()

We'll use the tapply command.  See https://rpubs.com/brouwern/groupeddata for more deetails

Calculate means and standard deviations for the brook trout present and brook trout absent rows of data.

NOTE: we'll using the original **salmon** dataframe, NOT the subsets


```{r}
#use tapply() on salmon dataframe
my.means <- tapply(salmon$percent.surv,
                       salmon$brook.trout.PRES.ABS,
                       FUN = mean,
                       na.rm = T)

#the result: both means calcualted
my.means

#use tapply() to calcualte the sd
my.sd <- tapply(salmon$percent.surv,
                       salmon$brook.trout.PRES.ABS,
                       FUN = sd,
                       na.rm = T)

#the result
my.sd

```



## Calculate standard error from the sd

```{r}
#sample size
n<- 6

#calculate se
my.se <- sqrt(my.sd)/n


```



# Plotting means

Boxplot and histograms are good for exploring the oveall distribution of the data.  We typically want to plot the mans with error bars when we are doing hypothesis testing.



## Load the plot2means() function

From https://rpubs.com/brouwern/plot2means
```{r}
#### The function STARTS here ####
plot2means <- function(means,
                           se,
                           groups = 2,
                           categories = c("Group 1","Group 2"),
                           x.axis.label = "x axis",
                           y.axis.label = "y axis",
                       axis.adjust = 0){
  
  # reset plot window
  par(mfrow = c(1,1), 
      mar = c(3.5,3,2,1))
                             
  # calculate values for plotting limits                 
  y.max <- max(means+2*se) +                    
    max(means+2*se)*axis.adjust
  y.min <- min(means-2*se) - 
    max(means+2*se)*axis.adjust
  
  if(groups == 2){ x.values <- c(0.25, 0.5)}
  
  
  
  #Plot means
  plot(means ~ x.values,
       xlim = c(0.2,0.55),
       ylim = c(y.min,y.max),
       xaxt = "n",
       xlab = "",
       ylab = "",
       cex = 1.25,
       pch = 16)
  
  axis(side = 1, 
       at = x.values,
       labels = categories,
      )
  
  #Plot upper error bar 
  lwd. <- 2
  arrows(y0 = means,
         x0 = x.values,
         y1 = means+2*se,
         x1 = x.values,
         length = 0,
         lwd = lwd.)
  
  #Plot lower error bar
  arrows(y0 = means,
         x0 = x.values,
         y1 = means-2*se,
         x1 = x.values,
         length = 0,
         lwd = lwd.) 
  
  mtext(text = x.axis.label,side = 1,line = 1.75)
  mtext(text = y.axis.label,side = 2,line = 1.95)
  mtext(text = "Error bars = 95% CI",side = 3,line = 0,adj = 0)
  
}

#### The function ENDS here ####
```




## Plot the means & the se
```{r}
plot2means(means = my.means,
           se = my.se,
           categories =  c("absent",
                           "present"),
           x.axis.label = "Brook trout status",
           y.axis.label = "Salmon survival"
            
          )
```



# Conducting a 2-sample t-test

We'll use the main salmon dataframe

```{r}
t.test(percent.surv ~ brook.trout.PRES.ABS, 
       data = salmon)
```


```{r, echo = F}
t.out <- t.test(percent.surv ~ brook.trout.PRES.ABS, 
       data = salmon)
mean1 <- round(t.out$estimate[[1]],3)
mean2 <- round(t.out$estimate[[2]],3)
```


Note that the output reports an interesting thing, the 95% confidence interval.  Specifically, this is the 95% for the difference between the two mean values.  The means are `r mean1` and  `r mean2`, and their difference is `r mean1 - mean2`.  

The ***null hypothesis (Ho)** is that this value is essentially equal to zero.  `r mean1 - mean2` is close to zero, and we use the t-test and confidence intervals to determine if a difference of `r mean1 - mean2` could just result from random sampling error.

If we plot this difference and the confidence interval we get the graph below.  The red line indicates where zero is on the y axis.

```{r,echo = F, warning=F, include=FALSE }
library(Hmisc)
par(mfrow = c(1,1), mar = c(4,4,4,4))
errbar(x = 1, y = mean1-mean2,
       yplus = t.out$conf.int[[2]],
       yminus = t.out$conf.int[[1]],
       xaxt = "n",ylab = "Diff. btwn means",xla = "")

abline(h=0, col = 2, lwd = 2,lty =2)
```



### Compare original mean values to "differene between the means"

```{r}
par(mfrow = c(1,2), mar = c(4,4,4,4))
errbar(x = 1:2, 
       y = my.means,
       yplus = my.means + 2*my.se,
       yminus = my.means -2*my.se,
       xaxt = "n",ylab = "Survival",
       xlim = c(0.75,2.25),
       main = "Original means")
mtext("NOTE: y axis is from 0.15 to 0.35",side = 1, line = 2)
axis(side = 1,at = c(1,2),labels = c("absent","present"))
mtext("Plot of means of each group")

errbar(x = 1, y = mean1-mean2,
       yplus = t.out$conf.int[[2]],
       yminus = t.out$conf.int[[1]],
       xaxt = "n",ylab = "Diff. btwn survival rates",
       main = "Difference between means",
       xlab = "")

abline(h=0, col = 2, lwd = 2,lty =2)
mtext("NOTE: y axis is from -0.05 to 0.15",side = 1, line = 2)
mtext("Plot difference between means")


```

