---
title: 'Lab 9: All Data'
author: "brouwern@gmail.com"
date: "November 6, 2016"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


```{r, echo = F}
# means = 3 mean values calcualted from raw data
# se = 3 standard errors
# categories = 3 categorical / factors that group the data
# x.axis.label = what should be plotted on the x axis
# y.axis.label = what should be plotted on the y axis


#### The function STARTS here ####
plot3means <- function(means,
                           se,
                           groups = 3,
                           categories = c("Group 1","Group 2", "Group 3"),
                           x.axis.label = "x axis",
                           y.axis.label = "y axis",
                       axis.adjust = 0){
  
  # reset plot window
  #par(mfrow = c(1,1), 
  #    mar = c(3.5,3,2,1))
                             
  # calculate values for plotting limits            
  y.max <- max(means+2*se) +                    
    max(means+2*se)*axis.adjust
  y.min <- min(means-2*se) - 
    max(means+2*se)*axis.adjust
  
  if(groups == 2){ x.values <- c(0.25, 0.5)}
  if(groups == 3){ x.values <- c(0.25, 0.5,0.75)}
  
  x.axis.min <- min(x.values)-0.05
  x.axis.max <- max(x.values)+0.05
  
  x.limits <- c(x.axis.min,x.axis.max)
  
  #Plot means
  plot(means ~ x.values,
       xlim = x.limits,
       ylim = c(y.min,y.max),
       xaxt = "n",
       xlab = "",
       ylab = "",
       cex = 1.25,
       pch = 16)
  
  axis(side = 1, 
       at = x.values,
       labels = categories,
      )
  
  #Plot upper error bar 
  lwd. <- 2
  arrows(y0 = means,
         x0 = x.values,
         y1 = means+2*se,
         x1 = x.values,
         length = 0,
         lwd = lwd.)
  
  #Plot lower error bar
  arrows(y0 = means,
         x0 = x.values,
         y1 = means-2*se,
         x1 = x.values,
         length = 0,
         lwd = lwd.) 
  
  mtext(text = x.axis.label,side = 1,line = 1.75)
  mtext(text = y.axis.label,side = 2,line = 1.95)
  mtext(text = "Error bars = 95% CI",side = 3,line = 0,adj = 0)
  
print("This function plots approximate 95% confidence intervals based on standard error imput by the user.  DO not enter 95% CIs")
  
}

#### The function ENDS here ####
```



```{r}
# means = 3 mean values calcualted from raw data
# se = 3 standard errors
# categories = 3 categorical / factors that group the data
# x.axis.label = what should be plotted on the x axis
# y.axis.label = what should be plotted on the y axis


#### The function STARTS here ####
plotTukeysHSD <- function(tukey.out,
                           x.axis.label = "Comparison",
                           y.axis.label = "Effect Size",
                       axis.adjust = 0){
  
  tukey.out <- as.data.frame(tukey.out[[1]])
  means <- tukey.out$diff
  categories <- row.names(tukey.out)
  groups <- length(categories)
  ci.low <- tukey.out$lwr
  ci.up  <- tukey.out$upr                         
   
  # reset plot window
  #par(mfrow = c(1,1), 
  #    mar = c(3.5,3,2,1))
                             
  # calculate values for plotting limits            
  y.max <- max(ci.up) +                    
    max(ci.up)*axis.adjust
  y.min <- min(ci.low) - 
    max(ci.low)*axis.adjust
  
  if(groups == 2){ x.values <- c(0.25, 0.5)}
  if(groups == 3){ x.values <- c(0.25, 0.5,0.75)}
  
  x.axis.min <- min(x.values)-0.05
  x.axis.max <- max(x.values)+0.05
  
  x.limits <- c(x.axis.min,x.axis.max)
  
  #Plot means
  plot(means ~ x.values,
       xlim = x.limits,
       ylim = c(y.min,y.max),
       xaxt = "n",
       xlab = "",
       ylab = "",
       cex = 1.25,
       pch = 16)
  
  axis(side = 1, 
       at = x.values,
       labels = categories,
      )
  
  #Plot upper error bar 
  lwd. <- 2
  arrows(y0 = means,
         x0 = x.values,
         y1 = ci.up,
         x1 = x.values,
         length = 0,
         lwd = lwd.)
  
  #Plot lower error bar
  arrows(y0 = means,
         x0 = x.values,
         y1 = ci.low,
         x1 = x.values,
         length = 0,
         lwd = lwd.) 
  
  mtext(text = x.axis.label,side = 1,line = 1.75)
  mtext(text = y.axis.label,side = 2,line = 1.95)
  mtext(text = "Error bars = 95% CI",side = 3,line = 0,adj = 0)
  
  
}
```



## Outline

* Make 4 datasets of different antler measurements (mass, basal circumference, main beam length, inside spread)
* ANOVA omnibus F-test
* Pairwise t-tests
* TukeyHSD tests
* Plot effect sizes from TukeyHSD



# Preliminaries: Make datasets

### Datset 1) Antler mass data

```{r}
mass<-c(472.5114,722.7774,548.6796,464.3168,577.2037,406.7964,555.8052,657.5226,734.7416,919.2983,665.6099,570.3627,429.5572,520.9726,346.0116,639.3096,544.8374,657.6512,802.2045,443.339,545.9901,791.3983,884.5803,763.2926,782.1005,840.2083,417.7051,735.8959,759.7498,877.0629)
n <- length(mass)/3
diet <- c(rep("Hi.Hi",n),rep("Hi.Lo",n),rep("Lo.Hi",n))
df.mass <- data.frame(antler.mass = mass,
                      diet = diet)
rm(list = c("mass", "diet","n"))
```

<br><br>

### Datset 2) Antler mass circumference data
```{r}
circum <- c(94.85394,83.21354,64.94166,96.9141,77.03692,113.2507,90.95746,95.83833,82.66041,97.46359,91.61469,96.4177,100.5873,99.3556,94.23969,57.94891,134.9324,95.01268,128.2977,129.307,70.39667,133.7504,121.0767,105.3898,86.81835,106.1906,136.0613,74.95854,112.6994,110.5714)


n <- length(circum)/3
diet <- c(rep("Hi.Hi",n),rep("Hi.Lo",n),rep("Lo.Hi",n))
df.circum <- data.frame(antler.circum = circum,
                        diet = diet)
rm(list = c("circum", "diet","n"))
```

<br><br>

### Datset 3)  Antler beam length data

```{r}
beam <- c(335.6464,410.2204,425.6838,355.1535,346.7697,399.8559,425.4827,447.6792,397.116,428.6749,406.034,380.342,437.5513,444.8916,437.9827,391.446,425.6301,404.2163,461.0517,568.163,457.6777,459.0168,270.5814,440.9724,412.0543,428.0725,247.2619,386.4118,423.5378,505.5221)
n <- length(beam)/3
diet <- c(rep("Hi.Hi",n),rep("Hi.Lo",n),rep("Lo.Hi",n))
df.beam <- data.frame(antler.beam = beam,
                      diet = diet)
rm(list = c("beam", "diet","n"))
```

<br><br>

### Datset 4) Antler spread data
```{r}
spread <- c(271.4356,291.3746,346.3026,345.6056,234.1441,339.7039,291.5156,275.9856,223.7703,316.9263,375.1109,384.1469,396.6894,272.4158,356.8185,393.6393,412.3185,425.7551,390.7663,381.4302,341.7604,350.6951,306.0166,354.8825,312.0478,459.7191,415.1186,321.8994,281.539,372.0439)
n <- length(spread)/3
diet <- c(rep("Hi.Hi",n),rep("Hi.Lo",n),rep("Lo.Hi",n))
df.spread <- data.frame(antler.spread = spread,
                        diet = diet)
rm(list = c("spread", "diet","n"))
```


# Data analysis: Mass data

### Summary stats for all mass data

#### Basic summary stats
```{r}
mean(df.mass$antler.mass)
var(mean(df.mass$antler.mass))
sd.mass <- sd(df.mass$antler.mass)
n.mass <- length(mean(df.mass$antler.mass))
```

#### Standard errors
```{r}
sd.mass/sqrt(n.mass)
```

<br><br>

### Summary stats by diet treatment

Use tapply to get means and standard deviations by group.

#### Calcualte means with tapply
```{r}
mean.mass <- tapply(df.mass$antler.mass,
                    df.mass$diet,
                    FUN = mean)
```

<br><br>

#### Calcualte sd with tapply
```{r}
sd.mass <- tapply(df.mass$antler.mass,
                    df.mass$diet,
                    FUN = sd)
```


#### Calcualte sample size with tapply
```{r}
n.mass <- tapply(df.mass$antler.mass,
                    df.mass$diet,
                    FUN = length)
```

#### Calcualte SD

```{r}
se.mass <- sd.mass/sqrt(n.mass)
```


### Plot the means

This requires the plot3means function from
https://rpubs.com/brouwern/plot3means

```{r}
categories <- levels(df.mass$diet)

plot3means(means = mean.mass,
           se = se.mass,
           categories = categories,
           y.axis.label = "Anterl mass (g)"
           )
```





<br><br>

### Plot raw data
```{r}
boxplot(antler.mass ~ diet,df.mass)
```

<br><br>

### ANOVA omnibus F-test

Build models
```{r}
model.null <- lm(antler.mass ~ 1, data = df.mass)
model.alt  <- lm(antler.mass ~ diet, data = df.mass)
```

Compare models to get F statistic and p-values
```{r}
anova(model.null, model.alt)
```

Save model output
```{r}
mass.anova.table <- anova(model.null, model.alt)
```


<br><br>

### Pairwise t-test

summary(df.mass)

Make table of pairwise t-test
```{r}
pairwise.t.test(x = df.mass$antler.mass,
                g = df.mass$diet,
                p.adjust.method = "none")
```


Save output
```{r}
mass.pairwise.p <- pairwise.t.test(x = df.mass$antler.mass,
                g = df.mass$diet,
                p.adjust.method = "none")
```


### TukeyHSD p-values

Fit model with aov() instead of lm()
```{r}
mass.aov <- aov(antler.mass ~ diet, data = df.mass)
```

Calcaulte corrected p valeus and effect sizes
```{r}
TukeyHSD(mass.aov)
```

<br><br><br>

### Plot effect sizes from TukeyHSD

Save output of TukeyHSD to an object
```{r}
mass.tuk <- TukeyHSD(mass.aov)
```

Use plotTukeyHSD from
https://rpubs.com/brouwern/plotTukeyHSD
```{r}
plotTukeysHSD(mass.tuk)
```

<br><br>


# Circumference data
`
# Data analysis

### Summary stats for all data

#### Basic summary stats
```{r}
mean(df.circum$antler.circum)
var(mean(df.circum$antler.circum))
sd.circum <- sd(df.circum$antler.circum)
n.circum <- length(mean(df.circum$antler.circum))
```

#### Standard errors
```{r}
sd.circum/sqrt(n.circum)
```

<br><br>

### Summary stats by diet treatment

Use tapply to get means and standard deviations by group.

#### Calcualte means with tapply
```{r}
mean.circum <- tapply(df.circum$antler.circum,
                    df.circum$diet,
                    FUN = mean)
```

<br><br>

#### Calcualte sd with tapply
```{r}
sd.circum <- tapply(df.circum$antler.circum,
                    df.circum$diet,
                    FUN = sd)
```


#### Calcualte sample size with tapply
```{r}
n.circum <- tapply(df.circum$antler.circum,
                    df.circum$diet,
                    FUN = length)
```

#### Calcualte SD

```{r}
se.circum <- sd.circum/sqrt(n.circum)
```


### Plot the means

This requires the plot3means function from
https://rpubs.com/brouwern/plot3means

```{r}
categories <- levels(df.circum$diet)

plot3means(means = mean.circum,
           se = se.circum,
           categories = categories,
           y.axis.label = "Anterl circum (g)"
           )
```





<br><br>

### Plot raw data
```{r}
boxplot(antler.circum ~ diet,df.circum)
```

<br><br>

### ANOVA omnibus F-test

Build models
```{r}
model.null <- lm(antler.circum ~ 1, data = df.circum)
model.alt  <- lm(antler.circum ~ diet, data = df.circum)
```

Compare models to get F statistic and p-values
```{r}
anova(model.null, model.alt)
```

Save model output
```{r}
circum.anova.table <- anova(model.null, model.alt)
```


<br><br>

### Pairwise t-test

summary(df.circum)

Make table of pairwise t-test
```{r}
pairwise.t.test(x = df.circum$antler.circum,
                g = df.circum$diet,
                p.adjust.method = "none")
```


Save output
```{r}
circum.pairwise.p <- pairwise.t.test(x = df.circum$antler.circum,
                g = df.circum$diet,
                p.adjust.method = "none")
```


### TukeyHSD p-values

Fit model with aov() instead of lm()
```{r}
circum.aov <- aov(antler.circum ~ diet, data = df.circum)
```

Calcaulte corrected p valeus and effect sizes
```{r}
TukeyHSD(circum.aov)
```

<br><br><br>

### Plot effect sizes from TukeyHSD

Save output of TukeyHSD to an object
```{r}
circum.tuk <- TukeyHSD(circum.aov)
```

Use plotTukeyHSD from
https://rpubs.com/brouwern/plotTukeyHSD
```{r}
plotTukeysHSD(circum.tuk)
```


# Antler main beam data

# Data analysis

### Summary stats for all data

#### Basic summary stats
```{r}
mean(df.beam$antler.beam)
var(mean(df.beam$antler.beam))
sd.beam <- sd(df.beam$antler.beam)
n.beam <- length(mean(df.beam$antler.beam))
```

#### Standard errors
```{r}
sd.beam/sqrt(n.beam)
```

<br><br>

### Summary stats by diet treatment

Use tapply to get means and standard deviations by group.

#### Calcualte means with tapply
```{r}
mean.beam <- tapply(df.beam$antler.beam,
                    df.beam$diet,
                    FUN = mean)
```

<br><br>

#### Calcualte sd with tapply
```{r}
sd.beam <- tapply(df.beam$antler.beam,
                    df.beam$diet,
                    FUN = sd)
```


#### Calcualte sample size with tapply
```{r}
n.beam <- tapply(df.beam$antler.beam,
                    df.beam$diet,
                    FUN = length)
```

#### Calcualte SD

```{r}
se.beam <- sd.beam/sqrt(n.beam)
```


### Plot the means

This requires the plot3means function from
https://rpubs.com/brouwern/plot3means

```{r}
categories <- levels(df.beam$diet)

plot3means(means = mean.beam,
           se = se.beam,
           categories = categories,
           y.axis.label = "Anterl beam (g)"
           )
```





<br><br>

### Plot raw data
```{r}
boxplot(antler.beam ~ diet,df.beam)
```

<br><br>

### ANOVA omnibus F-test

Build models
```{r}
model.null <- lm(antler.beam ~ 1, data = df.beam)
model.alt  <- lm(antler.beam ~ diet, data = df.beam)
```

Compare models to get F statistic and p-values
```{r}
anova(model.null, model.alt)
```

Save model output
```{r}
beam.anova.table <- anova(model.null, model.alt)
```


<br><br>

### Pairwise t-test

summary(df.beam)

Make table of pairwise t-test
```{r}
pairwise.t.test(x = df.beam$antler.beam,
                g = df.beam$diet,
                p.adjust.method = "none")
```


Save output
```{r}
beam.pairwise.p <- pairwise.t.test(x = df.beam$antler.beam,
                g = df.beam$diet,
                p.adjust.method = "none")
```


### TukeyHSD p-values

Fit model with aov() instead of lm()
```{r}
beam.aov <- aov(antler.beam ~ diet, data = df.beam)
```

Calcaulte corrected p valeus and effect sizes
```{r}
TukeyHSD(beam.aov)
```

<br><br><br>

### Plot effect sizes from TukeyHSD

Save output of TukeyHSD to an object
```{r}
beam.tuk <- TukeyHSD(beam.aov)
```

Use plotTukeyHSD from
https://rpubs.com/brouwern/plotTukeyHSD
```{r}
plotTukeysHSD(beam.tuk)
```


# Antler Spread Data

# Data analysis

### Summary stats for all data

#### Basic summary stats
```{r}
mean(df.spread$antler.spread)
var(mean(df.spread$antler.spread))
sd.spread <- sd(df.spread$antler.spread)
n.spread <- length(mean(df.spread$antler.spread))
```

#### Standard errors
```{r}
sd.spread/sqrt(n.spread)
```

<br><br>

### Summary stats by diet treatment

Use tapply to get means and standard deviations by group.

#### Calcualte means with tapply
```{r}
mean.spread <- tapply(df.spread$antler.spread,
                    df.spread$diet,
                    FUN = mean)
```

<br><br>

#### Calcualte sd with tapply
```{r}
sd.spread <- tapply(df.spread$antler.spread,
                    df.spread$diet,
                    FUN = sd)
```


#### Calcualte sample size with tapply
```{r}
n.spread <- tapply(df.spread$antler.spread,
                    df.spread$diet,
                    FUN = length)
```

#### Calcualte SD

```{r}
se.spread <- sd.spread/sqrt(n.spread)
```


### Plot the means

This requires the plot3means function from
https://rpubs.com/brouwern/plot3means

```{r}
par(mfrow = c(1,1))
categories <- levels(df.spread$diet)
plot3means(means = mean.spread,
           se = se.spread,
           categories = categories,
           x.axis.label = "Diet treatment",
           y.axis.label = "Anter spread (mm)",axis.adjust = 0.01
           )
```





<br><br>

### Plot raw data
```{r}
boxplot(antler.spread ~ diet,df.spread)
```

<br><br>

### ANOVA omnibus F-test

Build models
```{r}
model.null <- lm(antler.spread ~ 1, data = df.spread)
model.alt  <- lm(antler.spread ~ diet, data = df.spread)
```

Compare models to get F statistic and p-values
```{r}
anova(model.null, model.alt)
```

Save model output
```{r}
spread.anova.table <- anova(model.null, model.alt)
```


<br><br>

### Pairwise t-test

Make table of pairwise t-test
```{r}
pairwise.t.test(x = df.spread$antler.spread,
                g = df.spread$diet,
                p.adjust.method = "none")
```


Save output
```{r}
spread.pairwise.p <- pairwise.t.test(x = df.spread$antler.spread,
                g = df.spread$diet,
                p.adjust.method = "none")
```








### TukeyHSD p-values

Fit model with aov() instead of lm()
```{r}
spread.aov <- aov(antler.spread ~ diet, data = df.spread)
```

Calcaulte corrected p valeus and effect sizes
```{r}
TukeyHSD(spread.aov)
```

<br><br><br>

### Plot effect sizes from TukeyHSD

Save output of TukeyHSD to an object
```{r}
spread.tuk <- TukeyHSD(spread.aov)
```

Use plotTukeyHSD from
https://rpubs.com/brouwern/plotTukeyHSD
```{r}
plotTukeysHSD(spread.tuk)
```



<br><br>

All output

```{r}
mass.anova.table
mass.pairwise.p
mass.aov
mass.tuk

beam.anova.table
beam.pairwise.p
beam.aov
beam.tuk

circum.anova.table
circum.pairwise.p
circum.aov
circum.tuk

spread.anova.table
spread.pairwise.p
spread.aov
spread.tuk
```

