---
title: "Lab 9: Intro to ANOVA"
author: "brouwern@gmail.com"
date: "November 3, 2016"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Updates to this lab will be emailed and posted to the course Facebook page
https://www.facebook.com/groups/930301587096169/?ref=bookmarks

# Inspiration
Asleson et al. 1997.  Effects of seasonal protein restriction on antlerogenesis and body mass in adult male white-tailed deer.  Journal of Wildlife Management 61.

# Create data

Make dataframe for analysis
```{r}
mass<-c(472.5114,722.7774,548.6796,464.3168,577.2037,406.7964,555.8052,657.5226,734.7416,919.2983,665.6099,570.3627,429.5572,520.9726,346.0116,639.3096,544.8374,657.6512,802.2045,443.339,545.9901,791.3983,884.5803,763.2926,782.1005,840.2083,417.7051,735.8959,759.7498,877.0629)

n <- length(mass)/3

diet <- c(rep("Hi.Hi",n),rep("Hi.Lo",n),rep("Lo.Hi",n))

antlers <- data.frame(antler.mass = mass,
                 diet = diet)

rm(list = c("mass", "diet","n"))

```

Look at raw antler data
```{r}
summary(antlers)
```



# plot3means function
Need to load this function (or more general plotmeans() )
```{r, include=FALSE}
# means = 3 mean values calcualted from raw data
# se = 3 standard errors
# categories = 3 categorical / factors that group the data
# x.axis.label = what should be plotted on the x axis
# y.axis.label = what should be plotted on the y axis


#### The function STARTS here ####
plot3means <- function(means,
                           se,
                           groups = 3,
                           categories = c("Group 1","Group 2", "Group 3"),
                           x.axis.label = "x axis",
                           y.axis.label = "y axis",
                       axis.adjust = 0){
  
  # reset plot window
  #par(mfrow = c(1,1), 
  #    mar = c(3.5,3,2,1))
                             
  # calculate values for plotting limits            
  y.max <- max(means+2*se) +                    
    max(means+2*se)*axis.adjust
  y.min <- min(means-2*se) - 
    max(means+2*se)*axis.adjust
  
  if(groups == 2){ x.values <- c(0.25, 0.5)}
  if(groups == 3){ x.values <- c(0.25, 0.5,0.75)}
  
  x.axis.min <- min(x.values)-0.05
  x.axis.max <- max(x.values)+0.05
  
  x.limits <- c(x.axis.min,x.axis.max)
  
  #Plot means
  plot(means ~ x.values,
       xlim = x.limits,
       ylim = c(y.min,y.max),
       xaxt = "n",
       xlab = "",
       ylab = "",
       cex = 1.25,
       pch = 16)
  
  axis(side = 1, 
       at = x.values,
       labels = categories,
      )
  
  #Plot upper error bar 
  lwd. <- 2
  arrows(y0 = means,
         x0 = x.values,
         y1 = means+2*se,
         x1 = x.values,
         length = 0,
         lwd = lwd.)
  
  #Plot lower error bar
  arrows(y0 = means,
         x0 = x.values,
         y1 = means-2*se,
         x1 = x.values,
         length = 0,
         lwd = lwd.) 
  
  mtext(text = x.axis.label,side = 1,line = 1.75)
  mtext(text = y.axis.label,side = 2,line = 1.95)
  mtext(text = "Error bars = 95% CI",side = 3,line = 0,adj = 0)
  
print("This function plots approximate 95% confidence intervals based on standard error imput by the user.  DO not enter 95% CIs")
  
}

#### The function ENDS here ####
```



## plotTukeysHSD function
Need to load this function
```{r, include=FALSE}
# means = 3 mean values calcualted from raw data
# se = 3 standard errors
# categories = 3 categorical / factors that group the data
# x.axis.label = what should be plotted on the x axis
# y.axis.label = what should be plotted on the y axis


#### The function STARTS here ####
plotTukeysHSD <- function(tukey.out,
                           x.axis.label = "Comparison",
                           y.axis.label = "Effect Size",
                       axis.adjust = 0){
  
  tukey.out <- as.data.frame(tukey.out[[1]])
  means <- tukey.out$diff
  categories <- row.names(tukey.out)
  groups <- length(categories)
  ci.low <- tukey.out$lwr
  ci.up  <- tukey.out$upr                         
   
  # reset plot window
  #par(mfrow = c(1,1), 
  #    mar = c(3.5,3,2,1))
                             
  # calculate values for plotting limits            
  y.max <- max(ci.up) +                    
    max(ci.up)*axis.adjust
  y.min <- min(ci.low) - 
    max(ci.low)*axis.adjust
  
  if(groups == 2){ x.values <- c(0.25, 0.5)}
  if(groups == 3){ x.values <- c(0.25, 0.5,0.75)}
  
  x.axis.min <- min(x.values)-0.05
  x.axis.max <- max(x.values)+0.05
  
  x.limits <- c(x.axis.min,x.axis.max)
  
  #Plot means
  plot(means ~ x.values,
       xlim = x.limits,
       ylim = c(y.min,y.max),
       xaxt = "n",
       xlab = "",
       ylab = "",
       cex = 1.25,
       pch = 16)
  
  axis(side = 1, 
       at = x.values,
       labels = categories,
      )
  
  #Plot upper error bar 
  lwd. <- 2
  arrows(y0 = means,
         x0 = x.values,
         y1 = ci.up,
         x1 = x.values,
         length = 0,
         lwd = lwd.)
  
  #Plot lower error bar
  arrows(y0 = means,
         x0 = x.values,
         y1 = ci.low,
         x1 = x.values,
         length = 0,
         lwd = lwd.) 
  
  mtext(text = x.axis.label,side = 1,line = 1.75)
  mtext(text = y.axis.label,side = 2,line = 1.95)
  mtext(text = "Error bars = 95% CI",side = 3,line = 0,adj = 0)
  
  
}

#### The function ENDS here ####
```


<br><br>

# Part 1: Summary stats & plotting means
## 1) Summary stats on all data

In this section the grand mean on ALL the data is calculated; data is NOT broken up by treatment!  Data is considered by subgroup below

```{r}
# total sample size (all observations)
dim(antlers)

n.total <- length(antlers$antler.mass)

#mean of ALL samples
summary(antlers)

mean(antlers$antler.mass)

#variance of ALL samples
var(antlers$antler.mass)

#stdev of ALL samples
mass.sd <- sd(antlers$antler.mass)
```

<br><br>

## 2) SE for ** all ** data

This ignores treatments.  All data are combined / pooled.

```{r}
#Using saved sd
mass.se <- mass.sd/sqrt(n.total)

sqrt.n <- sqrt(n.total)

mass.se <- mass.sd/sqrt.n

#Using raw data
mass.se <- sd(antlers$antler.mass)/
             sqrt(length(antlers$antler.mass))
```

## 3) 95% CI for all data

This is the 95% confidence interval around the overall/grand mean.  

```{r}
1.96*mass.se
```


## 4) Calcualte summary stats for each group

Each group is considered seperately.
```{r}
# Means
mass.3.means <- tapply(antlers$antler.mass,
       antlers$diet,
       mean)

#SD
mass.3.sd <- tapply(antlers$antler.mass,
       antlers$diet,
       sd)

#sample size n
mass.3.n <- tapply(antlers$antler.mass,
       antlers$diet,
       length)
```


## 5) Calcualte the 3 SE values
```{r}

mass.3.se <- mass.3.sd/sqrt(mass.3.n)
```


## 6) Plot the 3 mean valeus

```{r}
#Using the levels command
antler.groups <- levels(antlers$diet)

#By hand
antler.groups <- c("Hi.Hi", "Hi.Lo", "Lo.Hi")

plot3means(means = mass.3.means,
           se = mass.3.se,
           categories = antler.groups,
           x.axis.label = "Diet treatment",
           y.axis.label = "Antler mass"
            )
```



```{r}
boxplot(antler.mass ~ diet, data = antlers)
```


<br><br><br>

# Part 2: Omnibus ANOVA test

## 7 & 8) Build the null (Ho) and alternative (Ha) models

```{r}
model.null <- lm(antler.mass ~ 1, 
                 data = antlers)

model.alt <- lm(antler.mass ~ diet, 
                 data = antlers)
```


<br><br>

## 9-12) Conduct the Omnibus test and get the ANOVA output

Produce the ANOVA table
```{r}
anova(model.null,
      model.alt)
```


## OPTIONAL:

If you are curious and things are going well, do the optional activities



### The summary command on an lm() object
```{r}
summary(model.null)
summary(model.alt)

```


### The anova command on a single lm() object
```{r}
anova(model.alt)
```


### Change how model is defined
```{r}
model.alt.2  <- lm(antler.mass ~-1 + diet, data = antlers)

summary(model.alt.2)
```


### Plotting model diagnostics
```{r}
plot(model.alt)
```



<br><br><br><br>

# Part 3: Pairwise comparisons

## 13-15) Get pairwise p-valeus
```{r}
pairwise.t.test(x = antlers$antler.mass,
                g = antlers$diet,
                p.adjust.method = "none")
```



<br><br><br>

# Part 4: Pairwise comparisons, correcting for multiple comparisons, & effect sizes


## 17) Build model with aov() function
```{r}
model.alt.aov <- aov(antler.mass ~ diet, 
                     data = antlers)


```

## 18) Get p-values with Tukey's HSD 

```{r}
TukeyHSD(model.alt.aov)
```




## 19) Plot effect sizes

```{r}
tukey.out <- TukeyHSD(model.alt.aov)

plotTukeysHSD(tukey.out)
abline(h = 0, col = 2, lty = 2)
```








## Make new data


#Make antler circumference data
```{r}
circum <- c(94.85394,83.21354,64.94166,96.9141,77.03692,113.2507,90.95746,95.83833,82.66041,97.46359,91.61469,96.4177,100.5873,99.3556,94.23969,57.94891,134.9324,95.01268,128.2977,129.307,70.39667,133.7504,121.0767,105.3898,86.81835,106.1906,136.0613,74.95854,112.6994,110.5714)


n <- length(circum)/3
diet <- c(rep("Hi.Hi",n),rep("Hi.Lo",n),rep("Lo.Hi",n))

df.circum <- data.frame(antler.circum = circum,
                 diet = diet)

rm(list = c("mass", "diet","n"))
df.circum
```








#Make beam length data
```{r}
beam <- c(335.6464,410.2204,425.6838,355.1535,346.7697,399.8559,425.4827,447.6792,397.116,428.6749,406.034,380.342,437.5513,444.8916,437.9827,391.446,425.6301,404.2163,461.0517,568.163,457.6777,459.0168,270.5814,440.9724,412.0543,428.0725,247.2619,386.4118,423.5378,505.5221)


n <- length(beam)/3
diet <- c(rep("Hi.Hi",n),rep("Hi.Lo",n),rep("Lo.Hi",n))

df.beam <- data.frame(antler.beam = beam,
                 diet = diet)

rm(list = c("beam", "diet","n"))
df.beam
```





### Make antler spread data
```{r}
spread <- c(271.4356,291.3746,346.3026,345.6056,234.1441,339.7039,291.5156,275.9856,223.7703,316.9263,375.1109,384.1469,396.6894,272.4158,356.8185,393.6393,412.3185,425.7551,390.7663,381.4302,341.7604,350.6951,306.0166,354.8825,312.0478,459.7191,415.1186,321.8994,281.539,372.0439)


n <- length(spread)/3
diet <- c(rep("Hi.Hi",n),rep("Hi.Lo",n),rep("Lo.Hi",n))

df.spread <- data.frame(antler.spread = spread,
                 diet = diet)

rm(list = c("spread", "diet","n"))
df.spread
```













## x) Subset data into eachseperate group

```{r}
df.Hi.Hi <- subset(antlers,
                   diet == "Hi.Hi")
df.Hi.Lo <- subset(antlers,
                   diet == "Hi.Lo")
df.Lo.Lo <- subset(antlers,
                   diet == "Lo.Lo")
```

## x)   Carry out a t-test on each combination

```{r}

t.test.HHvsHL <- t.test(df.Hi.Hi$antler.mass,
                        df.Hi.Lo$antler.mass)


t.test.HHvsHL <- t.test(df.Hi.Hi$antler.mass,
                        df.Hi.Lo$antler.mass)
```





