---
title: "Sample paired t-test"
author: "brouwern@gmail.com"
date: "October 27, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Paired T-test example


## Data
This data contains 6 paired observations.  There are six pairs of plots, each with an exclosure and a control treatment.  The exclosures are fenced to keep out deer, the controls are open and allow deer access.

This code builds up the dataframe we will use
```{r}
#LONG data - numeric data all in 1 column
plot <- c(1,1,3,3,4,4,5,5,6,6,7,7)
treatment <- rep(c("control","exclosure"),6)
species.richness <- c(3,3,6,5,2,3,3,1,3,2,5,2)
df.1column <- data.frame(plot = plot,
                 treatment = treatment,
                 species.richness = species.richness)

#reshape to wide data
library(reshape2)

df.2column <- dcast(data = df.1column,
                formula = plot ~ treatment,
                value.var = "species.richness",
                fun.aggregate = sum)
```


## Explore data
Explore the data using histograms

### Plot all data
Use the 1 column dataframe
```{r}
par(mfrow = c(1,1))
hist(df.1column$species.richness)
```
 

### Plot by group
Use the 2-column datafarme
```{r}
par(mfrow = c(1,2))
hist(df.2column$control, main = "control", xlab = "Species richness")
hist(df.2column$exclosure, main = "exclosure", xlab = "Species richness")
```

## Calculate the difference between treatments

Substract control from the exclosure
```{r}
df.2column$difference1 <- df.2column$exclosure - df.2column$control
```


Plot the differences
```{r}
par(mfrow = c(1,1))
hist(df.2column$difference1, main = "Difference: Excl - Control", xlab = "Difference")
```


# T test 1 on differences
```{r}
t.test(df.2column$difference1, mu = 0)
```




