---
title: "Transformations"
author: "brouwern@gmail.com"
date: "October 27, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a brief walk through of using log and square root transformations.  We'll first plot some data that is non-normal, transforme it, then look at the effects of transformation on the plots and p-values from t-tests.

# Functions used

* data.frame
* str
* par
* log
* sqrt
* hist
* t.test
* round


<br><br><br><br>

# Make the data

First, let's put some data into R objects
```{r}

diet.richness<- c(13, 22, 29, 22, 18, 21, 14, 14, 16, 12,25, 5,  9,  15, 10, 13, 11, 13, 21, 16,
6,  6,  15, 14, 13, 17, 15, 11, 11, 13,
11, 13, 12, 16, 11, 17, 18, 15, 13, 17,
15, 18, 15, 13, 12, 14, 12, 16, 15, 10,
21, 17, 15, 18)

stream <- c("POWD"	,"LAUREL"	,"LAUREL"	,"LAUREL"	,"LAUREL"	,"LAUREL"	,"LAUREL"	,"LAUREL"	,"LAUREL"	,"LAUREL"	,"LAUREL"	,"POWD","POWD","POWD"	,"POWD"	,"POWD"	,"POWD","POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD"	,"POWD")

```

<br><br><br><br>

Now, let's package up these data into a dataframe using the ** data.frame() ** command.
```{r}
df <- data.frame(stream = stream,
                 diet.richness = diet.richness)
```

<br><br><br><br>

## Look at the data using summary

```{r, echo = F}
summary(df)
```

<br><br><br><br>

## Look at the data using the st() "structure" command

ANother way to get a look at your data is using the str() command

```{r}
str(df)
```

<br>

This provides soome info about it.  I don't use this very often with dataframes, but the str() command comes in handy other times too.


<br><br><br><br>

## Change Margins

This isn't code you have to know, but I think it makes things easier for plotting.  HEre, I use the par() command to tweak the maring (mar = ) to be a little smaller.  THe defualts are c(5, 4, 4, 2) which results in some excessive white space around the entire graph (in my opinion).  
```{r}
par(mar = c(4,4,1.5,0.5))
```


<br><br><br><br>

## Make a histogram

What do you notice about this histogram?
```{r, echo = F}
par(mfrow = c(1,1))
hist(df$diet.richness, 
     xlab = "Diet Richness",
     main = "Distribution of number of diet items")
```

<br><br>

## Make boxplots by stream
```{r, echo = F}
boxplot(diet.richness ~ stream, 
     xlab = "Stream",
     ylab = "Diet Richness",
     main = "Diet items by stream")
```

<br><br>

## Transform the data

#### Log transformation

Use the ** log() ** command and the assignment operator ** <- ** to make a new column
```{r}
df$log.richness <- log(df$diet.richness)
```

<br><br>

#### Square root transformation

Use the ** sqrt() ** command to make a new column

```{r}
df$sqrt.richness <- sqrt(df$diet.richness)
```


<br><br>

## Make histogram of transformed data

### Log transformation

What do you notice about this histogram?

```{r, echo = F}
hist(df$log.richness, 
     xlab = "log(Diet Richness)",
     main = "Distribution of number of diet items")
```

<br><br>

### Square root transformation
What do you notice about this histogram?
```{r, echo = F}
hist(df$sqrt.richness, 
     xlab = "sqrt(Diet Richness)",
     main = "Distribution of number of diet items")
```

<br><br>

### Compare log and sqrt transformation with 2 histograms

This requires the command ** par(mfrow = c(1,2)) ** so that we get 2 plots side by side.
```{r}
par(mfrow = c(1,2))
hist(df$log.richness, 
     xlab = "log(Diet Richness)",
     main = "log transform")

hist(df$sqrt.richness, 
     xlab = "sqrt(Diet Richness)",
     main = "sqrt transform")
```

<br><br><br>


### Compare log, sqrt transformation AND original data with 3 histograms

This requires the command ** par(mfrow = c(1,3)) **.  NOte the ** c(1,3) **, which sets the plot for 1 row of plots and **3** columns.

```{r}
par(mfrow = c(1,3))
hist(df$diet.richness, 
     xlab = "Diet Richness",
     main = "raw data")
hist(df$log.richness, 
     xlab = "log(Diet Richness)",
     main = "log transform")

hist(df$sqrt.richness, 
     xlab = "sqrt(Diet Richness)",
     main = "sqrt transform")
```


<br><br><br><br><br><br>






## Make boxplots by stream of logged data
```{r, echo = F}
par(mfrow = c(1,1))
boxplot(log(diet.richness) ~ stream, 
     xlab = "Stream",
     ylab = "Diet Richness",
     main = "Diet items by stream")
```

<br><br><br><br>

## Make boxplots by stream of square root data
```{r, echo = F}
boxplot(sqrt(diet.richness) ~ stream, 
     xlab = "Stream",
     ylab = "Diet Richness",
     main = "Diet items by stream")
```

<br><br><br><br>


# Compare t-test on original w/raw data

### First t.test: raw data

NOTE: for the sake of this exercise, set **var.equal = T** .  Normally we would NOT do this
```{r}
t.test.1 <- t.test(diet.richness ~ stream,
                   data = df,
                   var.equal = T) #NOTE = normally we use the default of var.equal=F
t.test.1
```

<br><br>

### 2nd t.test: logged data
What code carries this out?
```{r, echo = F}
t.test.2 <- t.test(log(diet.richness) ~ stream,
                   data = df,var.equal = T)#NOTE = normally we use the default of var.equal=F
t.test.2
```

<br><br>

### 3rd t.test: sqrt data
What code carries this out?
```{r, echo = F}
t.test.3 <- t.test(sqrt(diet.richness) ~ stream,
                   data = df,
                   var.equal = T)#NOTE = normally we use the default of var.equal=F
t.test.3
```


<br><br>

### Examine results

Let's extract the p values.

A useful advanced comman in R is str().  It shows you the underlying structure and organization of R objects.

We can use it to see what is all packeged up in our t-test results.

```{r}
str(t.test.1)
```


<br>

This is all the raw info that gets formatted into a table by R from your t-test.  We can access individual parts of the output like this

```{r}
t.test.1$p.value
```

<br>

If we wanted the t-statistic we could do this

```{r}
t.test.1$statistic
```

<br><br>


Let's organize our 3 p values into a table to compare them.

```{r}
ps <- data.frame(p= c(t.test.1$p.value,
                      t.test.2$p.value,
                      t.test.3$p.value),
                 t= c(t.test.1$statistic,
                      t.test.2$statistic,
                      t.test.3$statistic),
                 test = c("raw","log","sqrt"))


```

<br><br>

Now look at the output
```{r}
ps
```

<br>

We don't really need this many decimal places so we can use the ** round() ** command to make thigns easier to read.  

```{r}
ps$p <- round(ps$p,5)
ps$t <- round(ps$t,2)
```

<br>
